\section{Design Details}
\label{sec:method}

\subsection{Problem Definition}

Given a protocol standard $S$ and an implementation $P$, \tool{} seeks to
identify whether $P$ satisfies a set of normative field-level requirements in
$S$. A requirement is represented as a rule
$r=\langle f, C, A, M, E\rangle$, where $f$ is the constrained protocol field,
$C$ is the applicability condition, $A$ is the expected action or constraint,
$M$ is the normative modality such as \emph{MUST} or \emph{SHOULD}, and $E$
records source evidence. The checker returns a tuple
$\langle y,E,U,T\rangle$: label $y\in\{\textsc{Consistent},
\textsc{Inconsistent},\textsc{Insufficient}\}$, evidence $E$, uncertainty notes
$U$, and optional validation-test artifact $T$.

\tool{} does not rely on naming similarity between protocol implementations and
standard text. Instead, it extracts a structured intermediate representation
from the standard and maps that representation to relevant variables and code
contexts in the implementation using a combination of semantic and syntactic
evidence.

\subsection{Field-Space Construction}
\label{sec:fieldspace}

Directly extracting rules from open-ended prose is unstable because standards
mix design motivation, compatibility advice, examples, and normative
requirements. \tool{} therefore starts by constructing a \emph{field space}: a
set of protocol objects that subsequent rules may reference. Field candidates
are collected from message formats, extension structures, parameter tables, and
explicit field-definition paragraphs. For each field, \tool{} records its
canonical name, aliases, parent message or structure, type, allowed values when
available, and source locations.

\textbf{Document preprocessing.}
\tool{} ingests RFC-style text, PDF, DOC, and TXT documents. To
preserve paragraph structure and local semantic coherence, it first splits the
document into paragraphs on blank lines and then greedily merges consecutive
paragraphs into text blocks. Paragraphs are never split during this process.
When adding the next paragraph would exceed a configurable maximum character
threshold, the current block is committed and a new block is started from that
paragraph. The default threshold is 6000 characters, and it can be adjusted
via a command-line argument. This character-level threshold provides stable,
tokenizer-independent block boundaries during preprocessing.

The field space serves two purposes. First, it gives the extractor a closed set
of referents, reducing the chance that descriptive concepts such as
``compatibility mode'' are mistaken for checkable fields. Second, it preserves
cross-document identity: the same field may appear as a full path, short name,
structure member, or contextual pronoun. During rule extraction, all references
must resolve to one or more field-space entries. Ambiguous references are kept
as uncertainty rather than silently collapsed.

\textbf{Field identity and validation.}
\tool{} treats two mentions as the same field if their parent context
and semantic role are compatible, preventing generic names such as
\emph{length}, \emph{type}, or \emph{flags} from being merged across unrelated
messages. Before rule extraction, it also validates the field space by marking
entries with missing parent context, incompatible aliases, or only weak prose
evidence as ambiguous or out of scope. This conservative policy favors
precision over recall because an incorrect field merge can mislead all
downstream alignment and reasoning.

\subsection{Field-Centric Rule Extraction}
\label{sec:rules}

After field-space construction, \tool{} uses each standard section as the basic
unit of analysis to extract field-centric rules from normative text.
Specifically, for each section, the system provides the section text, the
field-space entries relevant to that section, and a normative keyword lexicon
built from RFC 2119-style language to the model. Within these constraints, the
model identifies normative statements and resolves the protocol objects they
mention to concrete entries in the field space. It then emits structured
records under a fixed rule schema, including the constrained field,
applicability condition, expected action or constraint, normative modality,
and source evidence. \emph{MAY}-level statements are filtered out of this rule
set because they do not impose required behavior in our checking scope. The
resulting schema is summarized in Table~\ref{tab:ruleschema}.

\textbf{Local scope and validation.}
This design confines rule extraction to local sections and a known field set,
preventing open-ended inference over the full standard and reducing unsupported
rule generation. The field space provides a traceable referential basis for
protocol objects, so each rule can be linked back to specific fields and source
text. For normative sentences that contain multiple constraints, \tool{} splits
them into multiple field-level rules, each corresponding to an independent
checkable behavior such as field presence, value constraints, range checks,
state updates, or rejection behavior.

After extraction, \tool{} validates model outputs with a deterministic checker
to ensure structural validity and provenance traceability. The checks include
whether field references resolve to field-space entries, whether the source
span exists, whether the action type belongs to a supported action class, and
whether the rule duplicates an existing record. Outputs that fail validation
are retried with a concise repair prompt; if they still fail, the candidate
rule is discarded rather than turned into an unauditable rule.

\begin{table}[t]
\centering
\caption{Rule schema used by \tool.}
\label{tab:ruleschema}
\small
\begin{tabularx}{\columnwidth}{@{}l>{\raggedright\arraybackslash}X@{}}
\toprule
\textbf{Attribute} & \textbf{Description} \\
\midrule
Field & Constrained protocol field or parameter \\
Condition & Message type, state, phase, or field predicate \\
Action & Presence, value, derivation, relation, or rejection \\
Modality & MUST, MUST NOT, SHOULD, or equivalent phrase \\
Evidence & Source section, sentence span, and referenced text \\
\bottomrule
\end{tabularx}
\end{table}

The action families are intentionally close to implementation behavior:
\emph{assign}, \emph{derive}, \emph{select}, \emph{require-present},
\emph{require-absent}, \emph{compare}, \emph{range-check},
\emph{state-update}, and \emph{reject}. This design encourages a rule to map to
observable code constructs such as assignments, conditionals, error returns, and
state transitions. When a sentence contains multiple constraints, \tool{}
atomizes it into multiple rules so that each rule can be aligned and tested
independently.

\textbf{Rule normalization.}
Before alignment, rules are normalized to make equivalent requirements compare
more consistently. Negated forms are rewritten into explicit actions, e.g.,
``MUST NOT accept $x$'' becomes an expected \emph{reject} action under condition
$x$. Cross-references are expanded when the referenced section contains a
field-space entry. Multiple modalities in the same sentence are split into
separate rules. The normalized rule remains linked to the original text so that
reviewers can audit whether the transformation preserved the intended meaning.

\textbf{Unsupported rules.}
Some normative statements are intentionally not converted into field-level
rules. Examples include \emph{MAY}-level statements, broad performance
guidance, implementation advice without observable behavior, and cryptographic
assumptions that require a proof rather than code-level checking. \tool{}
records these statements as \emph{out-of-scope normative text}. Reporting
out-of-scope text is useful: it prevents silent loss of requirements and makes
the method's coverage explicit.

\subsection{Syntax-Aware Rule-to-Code Alignment}
\label{sec:alignment}

The alignment stage maps a rule $r$ to implementation variables and code
contexts. It proceeds in three steps.

\textbf{File ranking.}
\tool{} first builds a file-level summary for each candidate C/C++ file using
its path, function signatures, structure and enum definitions, comment summary,
and hits on field aliases. An LLM-based ranker then takes the rule, field
aliases, action class, and file summaries as input and estimates whether each
file is likely to contain the relevant field representation or checking logic.
The ranking considers field or alias matches, the relationship between file
functionality and the rule action, and the relationship between the file's
module and the target protocol object. The top-ranked files are deduplicated
and passed to variable localization.

\textbf{Variable localization.}
For the constrained field in the rule, \tool{} first constructs a set of field
aliases to bridge naming differences between standard field names and
implementation identifiers. The system tokenizes field names, normalizes naming
styles, and expands the alias set using common abbreviations, parent message
names, and parent structure names. It then builds a C/C++ symbol index with
Tree-sitter~\cite{tree_sitter}, covering variables, structure members, enum
constants, macros, and function parameters. For each indexed entity, \tool{}
records its declaration locations, use locations, scope, and enclosing
function.

On top of the symbol index, \tool{} ranks candidate identifiers using name
matches and syntactic usage patterns. Identifiers that exactly match a field
alias or have high token overlap receive higher priority. The rank is further
increased when the identifier appears in assignment, comparison, bounds
checking, state update, or error-return logic relevant to the rule action.
Conversely, identifiers with overly generic names, identifiers that appear only
as temporary or loop variables, and candidates without relevant syntactic usage
evidence are down-ranked. This stage outputs ranked implementation-variable
candidates and their declaration and use contexts as anchors for subsequent
code-context retrieval.

\textbf{Context retrieval.}
For each candidate variable $v$, \tool{} retrieves code snippets from three
sources: exact textual hits, symbol scopes containing the hits, and semantic
windows whose tokens overlap the rule. We use a fixed context-retrieval before 
ranking each snippet $c$ before LLM semantic filtering. The score is not intended to be
an optimal or learned ranking function. Instead, it instantiates an ordinal
retrieval-signal hierarchy: exact field or alias occurrence should dominate
variable-token overlap, variable-token overlap should dominate weak rule-token
overlap, and the action-role bonus should favor snippets that manipulate the
localized field in a way compatible with the rule action.
\begin{equation}
\begin{aligned}
S_{\mathrm{ctx}}(c)
&=
 w_{\mathrm{exact}}\cdot \mathbb{I}_{\mathrm{exact}}(c)
 +w_{\mathrm{var}}\cdot |\mathcal{T}_c \cap \mathcal{T}_v|\\
&\quad
 +w_{\mathrm{rule}}\cdot |\mathcal{T}_c \cap \mathcal{T}_r|
 +b(c).
\end{aligned}
\end{equation}
where $\mathbb{I}_{\mathrm{exact}}$ denotes exact variable or alias occurrence,
$\mathcal{T}_c$ is the code-token set, $\mathcal{T}_v$ is the variable-token set,
$\mathcal{T}_r$ is the rule-token set, and $b(c)$ rewards action-relevant usage
patterns such as initialization, validation, comparison, error handling, and
configuration reads. By default, $w_{\mathrm{exact}}=5$,
$w_{\mathrm{var}}=0.9$, and $w_{\mathrm{rule}}=0.28$. Snippets without exact
variable or alias occurrence receive a score cap to avoid weak semantic matches
dominating direct code matches.

The top snippets are then sent to an LLM semantic filter. The filter sees
only the rule and the ranked snippets, and selects the code contexts most
relevant to the rule. For each selected context, it explains the correspondence
between the snippet and the rule and returns a confidence score. Finally, the
model collects code contexts related to field representations, checking logic,
and error-handling paths for subsequent consistency reasoning.

The constants in Equation~(1) are therefore not a claim of universal
optimality. They encode a stable ordering among retrieval signals. Exact field
or alias occurrence is treated as the strongest signal because it directly links
the snippet to the localized implementation representation. Variable-token
overlap captures renamed or split representations. Rule-token overlap is weaker
because textual similarity to the standard often retrieves comments or unrelated
helper functions. The action-role bonus rewards snippets where the candidate field
participates in parsing, assignment, comparison, rejection, error propagation,
configuration reads, or state updates. We select the default constants once on
held-out sections and do not tune them using bug-finding outcomes.

\textbf{Why not pure retrieval?}
Pure semantic retrieval often retrieves a function whose comments resemble the
standard but whose variables are unrelated to the constrained field. Pure exact
matching has the opposite problem: it misses implementations that rename fields
or distribute a field across parser state and connection state. \tool{} uses
syntax-aware retrieval signals as the meeting point. A snippet is strongest
when it both mentions a plausible representation of the field and uses that
representation in a role consistent with the rule action.

\subsection{Consistency Reasoning}
\label{sec:reasoning}

After obtaining code contexts relevant to a rule, \tool{} enters the
consistency-reasoning stage. This stage confines reasoning to the current
evidence bundle rather than asking the model to inspect the entire codebase.
The bundle consists of the field rule, standard source span, localized
implementation variables, candidate code snippets, call relationships, and
configuration conditions. Based on this evidence, the reasoning agent decides
whether the observed implementation behavior satisfies the normative obligation
in the rule and outputs a structured result
$\langle y,E,U,T\rangle$. Here, $y$ is the judgment label and takes one of
\textsc{Consistent}, \textsc{Inconsistent}, or \textsc{Insufficient}; $E$
records the standard spans and code evidence supporting the judgment; $U$
records uncertainty in the current judgment and the missing context; and $T$
denotes validation-test artifacts, which are usually empty during static
reasoning and are populated later by the validation stage.

The reasoning process is evidence-bounded. The model may reason only from the
standard spans, code snippets, and call relationships in the evidence bundle,
and it may not assume behavior along code paths that have not been retrieved.
If the evidence shows that the implementation explicitly performs the check or
handling required by the rule, the result is \textsc{Consistent}. If the
evidence shows behavior that contradicts the normative requirement, the result
is \textsc{Inconsistent}. If the current evidence does not cover a key path,
such as a caller, macro definition, configuration branch, validation function,
or error-propagation path, the result is \textsc{Insufficient}. In that case,
$U$ records not only the reason for uncertainty but also the context request
needed for the next retrieval round, such as finding callers of a validation
function, the definition of a macro, the propagation path of an error code, or
the later uses of a field assignment.

\begin{algorithm}[t]
\small
\DontPrintSemicolon
\KwIn{Rule $r$, implementation index $I$, maximum rounds $K$}
\KwOut{Consistency tuple $\langle y,E,U,T\rangle$}
$C_0 \leftarrow$ AlignRuleToCode$(r,I)$\;
\For{$i \leftarrow 0$ \KwTo $K$}{
  $o_i \leftarrow$ Reason$(r,C_i)$\;
  \If{$o_i.y \neq \mathrm{Insufficient}$}{
    \Return $o_i$\;
  }
  $q_i \leftarrow$ ContextRequest$(o_i.U)$\;
  $C_{i+1} \leftarrow C_i \cup$ RetrieveAdditionalContext$(q_i,I)$\;
  \If{$C_{i+1}=C_i$}{
    \Return $o_i$\;
  }
}
\Return $o_K$\;
\caption{Evidence-driven consistency reasoning.}
\label{alg:reasoning}
\end{algorithm}

The system then supplements the code context according to the requests in $U$
and merges the new evidence into the bundle for another reasoning round.
Algorithm~\ref{alg:reasoning} describes this process. The loop terminates when
the reasoning agent returns a determinate label, when retrieval cannot provide
new useful context, or when the preset round limit is reached. Treating
\textsc{Insufficient} as an explicit result prevents \tool{} from mistaking
lack of evidence for implementation inconsistency. It also gives every
additional retrieval step a concrete target, improving the traceability and
reliability of the final consistency judgment.

\subsection{Validation-Test Agent}
\label{sec:testagent}

Candidate reports labeled \textsc{Inconsistent} by static reasoning may still
be false positives. A check that appears to be missing in a local snippet may
be enforced by an upper-level caller or state-machine path. Conversely, macro
definitions, build options, and conditional compilation can change which code
paths are active, making a judgment based only on local static evidence
incomplete or incorrect. \tool{} therefore introduces a validation-test agent
for potential inconsistencies, using executable behavior to further validate
the static reasoning result.

The validation-test agent receives the field rule, defect hypothesis, and
relevant code snippets, and generates a rule-directed test plan. The plan
attempts to construct two kinds of counterexamples: inputs that violate the
standard requirement but are accepted by the implementation, and inputs that
satisfy the standard requirement but are rejected by the implementation. The
agent then prepares the runtime environment, selects required build options,
generates or adapts a test harness, executes the test, and collects runtime
logs. 

\tool{} confirms a report as inconsistent only when the test stably
reaches the target path and the observed behavior contradicts the field rule.
If the test cannot reach the relevant path, or if the result is insufficient to
support the conclusion, the system records the unmet validation conditions
rather than promoting the static defect hypothesis to a confirmed issue. This
conservative policy ensures that confirmed results are grounded in reproducible
runtime evidence.

The validation runner supports three execution modes: extending an existing unit
test, constructing a small harness around a parser or validation function, and
driving a command-line implementation with generated inputs. For each run, the
system records compiler options, environment variables, execution commands,
exit status, stdout, stderr, and relevant logs. Build failures, unreachable
paths, and insufficient observations are classified as unmet validation
conditions and are not counted as confirmed inconsistencies.

After validation, \tool{} generates a structured validation report and
reproduction artifacts. The report includes the field rule, related standard
text, key code snippets, inconsistency rationale, test input, build
configuration, execution command, observed result, and log summary. For
reproducible inconsistencies, the report provides reproduction instructions and
script commands so maintainers can rerun the test under the same code version
and build conditions. For unconfirmed results, the report explicitly lists the
failure reason, such as build failure, unreachable target path, missing test
entry point, or insufficient observation signal. In this way, validation is not
only a confirmation step but also a traceable evidence chain for manual review
and maintainer communication.
