\section{Discussion and Limitations}
\label{sec:discussion}

\noindent \textbf{Checking scope.}
\tool{} primarily targets requirements that can be mapped to protocol fields,
trigger conditions, and implementation behavior. It is well suited for checking
local conformance deviations in parsing, validation, negotiation, and error
handling. For global properties such as cross-session behavior or long-term
state evolution, \tool{} can only provide limited approximations and cannot
offer completeness or correctness guarantees; strict conclusions still require
separate modeling and verification.

\noindent \textbf{Natural-language ambiguity.}
Ambiguity in standard text can affect rule extraction and conformance
judgment. For example, a requirement such as "reject malformed messages" may
not clearly specify the relevant field requirements or error-handling paths,
leading to false judgments or missed reports. For rules with unclear semantic
boundaries, the final conclusion still requires expert interpretation and
maintainer feedback; our artifact preserves the source text, reasoning
evidence, and uncertainty notes to support subsequent review.

\noindent \textbf{LLM dependence.}
\tool{} uses LLMs for extraction, ranking, semantic filtering, reasoning, and
test planning. The framework constrains these calls with schemas, field spaces,
syntax evidence, and deterministic validators, but it does not eliminate model
error. We mitigate this risk by retaining uncertainty, using low-temperature
decoding, validating outputs, and requiring executable evidence for confirmed 
reports.

\noindent \textbf{Cost and Human Effort.}
\tool{} reduces model cost by caching standard-extraction results,
compressing code contexts, and invoking the validation stage only for
suspicious samples or samples with insufficient evidence. Nevertheless,
multi-round reasoning and validation still incur higher model-calling costs
than one-shot prompting. Human effort mainly comes from pre-submission review
of candidate inconsistency reports, including checking report accuracy,
screening potential false positives, confirming execution results, and
submitting reports to maintainers with necessary follow-up communication.
Because each candidate report already contains the standard span, code
evidence, and validation result, human review usually does not require
reanalyzing the issue itself. On average, reviewing a candidate report takes
about two minutes. In addition, the number of misjudgments is small: among the
adjudicated reports, only four were rejected by maintainers as false positives.


\noindent \textbf{Source Code and Language Support.}
The current prototype relies on source code to locate variables, call
relationships, validation logic, and error-handling paths, and therefore does
not support closed-source implementations. \tool{} also depends on explicit
standard documents; if a standard is missing or does not sufficiently describe
field semantics and expected behavior, standard-to-implementation alignment is
limited. In addition, the current prototype mainly supports C/C++ protocol
implementations. For other programming languages, the system still requires
corresponding language front-ends that provide equivalent code-analysis
capabilities. In the future, we plan to extend this capability by integrating
front-end analysis components for additional languages.

\noindent \textbf{Result Adjudication.}
Protocol conformance judgments may be affected by ambiguity in standards and by
compatibility policies adopted by implementations. Therefore, system outputs
should not be treated as final adjudication results by themselves. We adjudicate
reports based on maintainer feedback, independent human review, and reproducible
tests, and classify them as valid issues, false positives, or pending. Thus, the
statistics reported in this paper reflect the adjudication status at the time of
evaluation, rather than the final lifecycle outcome of all submitted reports. In
addition, manually labeled gold standards may be influenced by annotators'
interpretations of ambiguous standard text, and LLM nondeterminism may change
rule extraction and reasoning results across runs. We mitigate these threats by
using two-author adjudication, preserving raw evidence, separating pending
reports from confirmed reports, and measuring variance across repeated GPT-5.4
runs and different model families.




% \noindent \textbf{Checking Scope and Ambiguity.}
% \tool{} primarily targets requirements that can be mapped to protocol fields, trigger conditions, and implementation behavior. It is well suited for local conformance deviations in parsing, validation, negotiation, and error handling. For global properties such as cross-session behavior, long-term state evolution, or requirements with ambiguous semantic boundaries, \tool{} can only provide limited conclusions; strict judgments still require expert interpretation, maintainer feedback, or separate modeling and verification. To support such review, the artifact preserves the source text, reasoning evidence, and uncertainty notes.

% \noindent \textbf{Implementation and Language Support.}
% The current prototype relies on source code to locate variables, call relationships, validation logic, and error-handling paths. It therefore does not support closed-source implementations, and its alignment quality may be affected by code generators, heavy macro use, or implementations whose semantic cues are incomplete in the source. The current front-end mainly supports C/C++ protocol implementations; other languages require corresponding analysis front-ends with equivalent capabilities.

% \noindent \textbf{LLM Dependence.}
% \tool{} uses LLMs for extraction, ranking, semantic filtering, reasoning, and test planning. Although these calls are constrained by schemas, field spaces, syntax evidence, and deterministic validators, model errors cannot be fully eliminated. We mitigate this risk by retaining uncertainty, using low-temperature decoding, validating outputs, and requiring executable evidence for confirmed reports.

% \noindent \textbf{Cost and Human Effort.}
% \tool{} reduces model cost by caching standard-extraction results, compressing code contexts, and invoking validation only for suspicious samples or samples with insufficient evidence. Multi-round reasoning and validation still incur higher model-calling costs than one-shot prompting, but the required human effort is limited. Because each candidate report already contains the standard span, code evidence, and validation result, pre-submission review usually does not require reanalyzing the issue itself. On average, reviewing a candidate report takes about two minutes; among the adjudicated reports, only four were rejected by maintainers as false positives.

% \noindent \textbf{Threats to Validity.}
% Our evaluation may be affected by manual adjudication, maintainer feedback, and model stability. Standards may be ambiguous, implementations may adopt compatibility-driven behavior, and some submitted reports remain pending due to external response latency. LLM nondeterminism may also change extraction or reasoning results across runs. We mitigate these threats by using two-author adjudication, preserving raw evidence, separating pending reports from confirmed reports, and measuring variance across repeated GPT-5.4 runs and different model families.
