\section{Conclusion}
\label{sec:conclusion}

This paper studied how to identify field-level conformance deviations in source
code implementations from natural-language protocol standards. We presented
\tool{}, which uses protocol fields as the linking points between standard text
and implementation code, extracts structured rules from standards, locates
relevant code evidence, and produces auditable conformance judgments. \tool{}
requires each judgment to be traceable to concrete standard spans, code
contexts, and validation results, thereby reducing the risk of misclassification.
Across multiple real protocol implementations, we submitted 204 reports, of
which 126 were confirmed or fixed by maintainers, and discovered two previously
unknown CVEs. These results show that starting from field requirements and
combining them with code evidence is an effective way to uncover standard
deviations in protocol implementations.



\noindent \textbf{Data Availability.}
Our evaluation is designed to support reviewer inspection without violating
double-blind review. For reports that have been confirmed and fixed by
maintainers, the artifact uses blinded identifiers and provides the normative
standard span, implementation evidence, inconsistency explanation, and
reproduction scripts. For reports that are still pending or may involve
undisclosed risks, we do not provide full reproduction materials in order to
avoid amplifying potential security impact. All code, scripts, and blinded
reports are included in an anonymous GitHub artifact repository submitted with
the paper:
\url{https://anonymous.4open.science/r/SpecField-Artifact}.
